Functional Pearl: Short and Mechanized Logical Relation for Dependent Type Theories

### 2024-02-28 paper/paper.tex
Revert I-Pi to use a function

\Rref{I-Pi} is the most complicated rule in our logical
relation. Instead of explaining it directly, we consider the following
simplified admissible \rref{I-PiAlt} that follows directly from \rref{I-Pi}.
\begin{center}
  \drule[]{I-PiAlt}
\end{center}
\begin{lemma}[I-PiAlt admissibility]
  \label{lemma:piintroalt}
  \Rref{I-PiAlt} is admissible.
\end{lemma}
\begin{proof}
Immediate from \rref{I-Pi} with $[[R]]$ instantiated to the relation $[[{ (a
, S0 ) | a in S , Interp I i B { a /x } S0 }]]$.
\end{proof}
Compared to the \rref{I-Pi}, which contains the mysterious relation
$[[R]]$, \rref{I-PiAlt} directly captures the meaning of
a well-behaved dependent function type. The precondition of
\rref{I-PiAlt} says that
the function type $[[Pi x : A . B]]$ has an interpretation if its input
type $[[A]]$ can be
interpreted as some set $[[S]]$, and for all terms $[[a in S]]$, the
type $[[B {a / x}]]$, obtained by substituting $[[a]]$ into the output type
$[[B]]$, has some semantic interpretation. In its conclusion, we say
that a term $[[b]]$ belongs to an interpretation of $[[Pi x : A . B]]$ if
for all $[[a in S]]$, where $[[S]]$ is an interpretation $[[A]]$, the
application form $[[b a]]$ belongs to all possible interpretations of
$[[B {a  / x}]]$ (the pre-condition ensures at least one interpretation
exists for each $[[B {a / x}]]$ where $[[a in S]]$).

In fact, while \rref{I-PiAlt} is an instantiation of \rref{I-Pi},
these two rules are equivalent in the sense that every derivation of
involving \rref{I-Pi} can be systematically replaced by
\rref{I-PiAlt}. This equivalence follows directly from the fact
that the logical relation is a partial function, a result we will show
in Lemma~\ref{lemma:logreldeter}. The preconditions of \rref{I-Pi},
when combined with the functionality of the logical relation, uniquely
determine the relation $[[R in S * PowerSet STm]]$ to be the relation
$[[{ (a , S0 ) | a in S , Interp I  i B { a /x } S0 }]]$. This result is
formally shown through the improved inversion lemma for function types
(Lemma~\ref{lemma:piinvalt}).

Unfortunately, we cannot define the function case of our logical
relation directly using \rref{I-PiAlt} since the occurrence of
$[[Interp I i B {a/x} S0]]$ in its conclusion not only violates the
syntactic strict positivity constraint required in proof assistants,
but is genuinely non-monotone when we rewrite the inductive definition
as an endofunction over the domain of relations.
Intuitively, the failure of monotonicity stems from the fact
that the witness picked in the precondition is not necessarily the
same witness being referred to in the post condition as the relation
grows, whereas the
relation $[[R]]$ in \rref{I-Pi} ``fixes'' the witnesses $[[S0]]$ and
prevents the set of witnesses from growing. While it might
be possible to restrict the domain with additional constraints such as
functionality and inversion properties to justify the well-definedness of our
inductive relation with \rref{I-PiAlt}, we opt for our current
formulation of \rref{I-Pi} so we immediately obtain a
well-defined inductive relation and a usable induction principle. The
slight disadvantage of \rref{I-Pi} is that we need to construct the
function $[[R]]$ each time we apply it, though this is mitigated by
the immediate admissibility of \rref{I-PiAlt} and the alternative $\Pi$ inversion
principle (Lemma~\ref{lemma:piinvalt}), which we derive after proving functionality.

@@

First, we prove a family of simple properties, which we refer to as
inversion principles for our logical relation. Given $[[Interp I i A
S]]$ where $[[A]]$ is in some head form such as $[[Bool]]$ or $[[Pi x
: A0 .B0]]$, the inversion lemma allows us to say something about the set
$[[S]]$. Its proof is simple, but we sketch out the case for
functions to help readers confirm their understanding of \rref{I-Pi}.
\begin{lemma}[Inversion of the logical relation]
  \label{lemma:interpinv}\leavevmode
  \begin{enumerate}
  \item\footnote{\dotv{semtyping.v}{InterpExt\_Void\_inv}} If $[[Interp I i Void S]]$, then $[[S = emptyset]]$.
  \item\footnote{\dotv{semtyping.v}{InterpExt\_Bool\_inv}} If $[[Interp I i Bool S]]$, then $[[S = { a | a =>+ true \/ a =>+ false   }]]$.
  \item\footnote{\dotv{semtyping.v}{InterpExt\_Eq\_inv}} If $[[Interp I i a ~ b : A S]]$, then $[[S = { p | p =>+ refl , a <=> b  }]]$.
  \item\footnote{\dotv{semtyping.v}{InterpExt\_Fun\_inv}} If $[[Interp
    I i Pi x : A . B S1]]$, then there exists $[[S]],[[R]]$ such that:
    \begin{itemize}
    \item $[[Interp I i A S ]]$
    \item $[[R in S * PowerSet STm]]$
    \item $[[forall a, (# a in S implies (# exists S0 , (a , S0) in R #) #)]]$
    \item $[[forall a, forall S0, (# ( a , S0 ) in R implies Interp I
      i B { a /x } S0 #)]]$
    \item $[[S1 = { b | forall a, forall S0, (# ( a , S0 ) in R implies b a in S0 #) }]]$
    \end{itemize}
  \item\footnote{\dotv{semtyping.v}{InterpExt\_Univ\_inv}} If $[[Interp I i Set j S]]$, then $[[j < i]]$ and $[[S = I j]]$.
  \end{enumerate}
\end{lemma}
\begin{proof}
  As mentioned earlier, we only show the inversion property for the
  function type.
  We start by inducting over the derivation of $[[Interp I i Pi x : A . B S]]$. There
  are only two possible cases we need to consider.
  \begin{description}
  \item[\Rref{I-Pi}:] Immediate.
  \item[\Rref{I-Red}:] We are given that $[[Interp I i Pi x : A . B S1]]$.
    We know that there exists some $[[A0]]$ and
    $[[B0]]$ such that $[[Pi x : A . B => Pi x : A0 . B0]]$ and $[[Interp I i Pi
    x : A0 . B0 S1]]$. From the
    induction hypothesis, there exists $[[S]]$ and $[[R]]$ such that :
    \begin{itemize}
    \item $[[Interp I i A0 S ]]$
    \item $[[R in S * PowerSet STm]]$
    \item $[[forall a, (# a in S implies (# exists S0 , (a , S0) in R #) #)]]$
    \item $[[forall a, forall S0, (# ( a , S0 ) in R implies Interp I
      i B0 { a / x } S0 #)]]$
    \item $[[S1 = { b | forall a, forall S0, (# ( a , S0 ) in R implies b a in S0 #) }]]$
    \end{itemize}
    By inverting the derivation of $[[Pi x : A . B => Pi x : A0 . B0]]$, we derive $[[A => A0]]$ and
    $[[B => B0]]$. By Lemma~\ref{lemma:parsubst}, we have $[[B {a /x} => B0 {a/x} ]]$ for all
    $[[a]]$. As a result, by \rref{I-Red}, the same $[[S]]$ and
    $[[R]]$ additionally satisfies the following properties:
    \begin{itemize}
    \item $[[Interp I i A S ]]$
    \item $[[forall a, forall S0, (# ( a , S0 ) in R implies Interp I
      i B { a /x } S0 #)]]$
    \end{itemize}
    These properties are exactly what we need to finish the proof.
  \end{description}
\end{proof}

@@

Lemma~\ref{lemma:logreldeter} enables us to show the following
improved inversion lemma for function types whose statement is free of
the relation $[[R]]$, analogous to the admissible rule \rref{I-PiAlt}.
\begin{lemma}[Pi Inversion Alt\footnote{\dotv{semtyping.v}{InterpExt\_Fun\_inv\_nopf}}]
  \label{lemma:piinvalt}
  Suppose $[[Interp I i Pi x : A . B S]]$, then there exists some $[[S0]]$
  such that the following constraints hold:
  \begin{itemize}
  \item $[[Interp I i A S0]]$
  \item $[[forall a, (# a in S0 implies (# exists S1 , Interp I i B {a
    / x}
    S1 #) #)]]$
  \item $[[S = { b | forall a, (# a in S0 , forall
    S1, (# Interp I i B {a / x} S1,  b a in S1 #) #) }]]$
  \end{itemize}
\end{lemma}
\begin{proof}
  Immediate from Lemmas~\ref{lemma:interpinv} and \ref{lemma:logreldeter}.
\end{proof}

@@

% The inductive
% definition of the logical relation in Figure~\ref{fig:logrel} requires
% the impredicativity of Coq's \texttt{Prop} sort since\scw{or ``so that''?} in \rref{I-Pi},
% the function $[[F]]$ can be later instantiated into the logical
% relation itself (e.g. in the proof of Lemma~\ref{lemma:piintroalt}).

>>>>

\Rref{I-Pi} is the most complicated rule in our logical
relation. Instead of explaining it directly, we consider the following
simplified admissible \rref{I-PiAlt} that follows directly from \rref{I-Pi}.
\begin{center}
  \drule[]{I-PiAlt}
\end{center}
\begin{lemma}[I-PiAlt admissibility]
  \label{lemma:piintroalt}
  \Rref{I-PiAlt} is admissible.
\end{lemma}
\begin{proof}
The precondition $[[forall a, (# a in S
implies (# exists S0 , Interp I i B { a / x } S0 #) #)]]$ from
\rref{I-PiAlt} immediately induces a function $[[F in S -> PowerSet
STm]]$ such that $[[forall a, (# a in S implies Interp I i B { a / x }
F a #)]]$, which is exactly what we need to apply \rref{I-Pi}.
\end{proof}
\Rref{I-PiAlt} directly captures the meaning of
a well-behaved dependent function type. The precondition of
\rref{I-PiAlt} says that
the function type $[[Pi x : A . B]]$ has an interpretation if its input
type $[[A]]$ can be
interpreted as some set $[[S]]$, and for all terms $[[a in S]]$, the
type $[[B {a / x}]]$, obtained by substituting $[[a]]$ into the output type
$[[B]]$, has some semantic interpretation. In its conclusion, we say
that a term $[[b]]$ belongs to an interpretation of $[[Pi x : A . B]]$ if
for all $[[a in S]]$, where $[[S]]$ is an interpretation $[[A]]$, the
application form $[[b a]]$ belongs to all possible interpretations of
$[[B {a  / x}]]$ (the pre-condition ensures at least one interpretation
exists for each $[[B {a / x}]]$ where $[[a in S]]$).

In fact, while \rref{I-PiAlt} is an instantiation of \rref{I-Pi},
these two rules are equivalent in the sense that every derivation of
involving \rref{I-Pi} can be systematically replaced by
\rref{I-PiAlt}. This equivalence follows directly from the fact
that the logical relation is a partial function, a result we will show
in Lemma~\ref{lemma:logreldeter}. The preconditions of \rref{I-Pi},
when combined with the functionality of the logical relation, uniquely
determine the function $[[F in S -> PowerSet STm]]$ to be the
functional relation
$[[{ (a , S0 ) | a in S , Interp I  i B { a /x } S0 }]]$. This result is
formally shown through the improved inversion lemma for function types
(Lemma~\ref{lemma:piinvalt}).

Unfortunately, we cannot define the function case of our logical
relation directly using \rref{I-PiAlt} since the occurrence of
$[[Interp I i B {a/x} S0]]$ in its conclusion not only violates the
syntactic strict positivity constraint required in proof assistants,
but is genuinely non-monotone when we treat the inductive definition
as the fixed point of an endofunction over the domain of relations.
Intuitively, the failure of monotonicity stems from the fact
that the witness picked in the precondition is not necessarily the
same witness being referred to in the post condition as the relation
grows, whereas the
function $[[F]]$ in \rref{I-Pi} ``fixes'' the witnesses $[[S0]]$ as
$[[F a]]$ for each $[[a in S]]$,
thus preventing the set of witnesses from growing. While it might
be possible to restrict the domain with additional constraints such as
functionality and inversion properties to justify the well-definedness of our
inductive relation with \rref{I-PiAlt}, we opt for our current
formulation of \rref{I-Pi} so we immediately obtain a
well-defined inductive relation and a usable induction principle. The
slight disadvantage of \rref{I-Pi} is that we need to construct the
function $[[F]]$ each time we apply it, though this is mitigated by
the immediate admissibility of \rref{I-PiAlt} and the alternative $\Pi$ inversion
principle (Lemma~\ref{lemma:piinvalt}), which we derive after proving functionality.

@@

First, we prove a family of simple properties, which we refer to as
inversion principles for our logical relation. Given $[[Interp I i A
S]]$ where $[[A]]$ is in some head form such as $[[Bool]]$ or $[[Pi x
: A0 .B0]]$, the inversion lemma allows us to say something about the set
$[[S]]$. Its proof is simple, but we sketch out the case for
functions to help readers confirm their understanding of \rref{I-Pi}.
\begin{lemma}[Inversion of the logical relation]
  \label{lemma:interpinv}\leavevmode
  \begin{enumerate}
  \item\footnote{\dotv{semtyping.v}{InterpExt\_Void\_inv}} If $[[Interp I i Void S]]$, then $[[S = emptyset]]$.
  \item\footnote{\dotv{semtyping.v}{InterpExt\_Bool\_inv}} If $[[Interp I i Bool S]]$, then $[[S = { a | a =>+ true \/ a =>+ false   }]]$.
  \item\footnote{\dotv{semtyping.v}{InterpExt\_Eq\_inv}} If $[[Interp I i a ~ b : A S]]$, then $[[S = { p | p =>+ refl , a <=> b  }]]$.
  \item\footnote{\dotv{semtyping.v}{InterpExt\_Fun\_inv}} If $[[Interp
    I i Pi x : A . B S1]]$, then there exists $[[S]],[[F]]$ such that:
    \begin{itemize}
    \item $[[Interp I i A S ]]$
    \item $[[F in S -> PowerSet STm]]$
    \item $[[forall a, (# a in S implies Interp I i B { a / x } F a #)]]$
    \item $[[S1 = { b | forall a, (# a in S implies b a in F a #) }]]$
    \end{itemize}
  \item\footnote{\dotv{semtyping.v}{InterpExt\_Univ\_inv}} If $[[Interp I i Set j S]]$, then $[[j < i]]$ and $[[S = I j]]$.
  \end{enumerate}
\end{lemma}
\begin{proof}
  As mentioned earlier, we only show the inversion property for the
  function type.
  We start by inducting over the derivation of $[[Interp I i Pi x : A . B S]]$. There
  are only two possible cases we need to consider.
  \begin{description}
  \item[\Rref{I-Pi}:] Immediate.
  \item[\Rref{I-Red}:] We are given that $[[Interp I i Pi x : A . B S1]]$.
    We know that there exists some $[[A0]]$ and
    $[[B0]]$ such that $[[Pi x : A . B => Pi x : A0 . B0]]$ and $[[Interp I i Pi
    x : A0 . B0 S1]]$. From the
    induction hypothesis, there exists $[[S]]$ and $[[F]]$ such that :
    \begin{itemize}
    \item $[[Interp I i A0 S ]]$
    \item $[[F in S -> PowerSet STm]]$
    \item $[[forall a, (# a in S implies Interp I i B0 { a / x } F a #)]]$
    \item $[[S1 = { b | forall a, (# a in S implies b a in F a #) }]]$
    \end{itemize}
    By inverting the derivation of $[[Pi x : A . B => Pi x : A0 . B0]]$, we derive $[[A => A0]]$ and
    $[[B => B0]]$. By Lemma~\ref{lemma:parsubst}, we have $[[B {a /x} => B0 {a/x} ]]$ for all
    $[[a]]$. As a result, by \rref{I-Red}, the same $[[S]]$ and
    $[[F]]$ additionally satisfies the following properties:
    \begin{itemize}
    \item $[[Interp I i A S ]]$
    \item $[[forall a, (# a in S implies Interp I i B { a / x } F a #)]]$
    \end{itemize}
    These properties are exactly what we need to finish the proof.
  \end{description}
\end{proof}

@@

Lemma~\ref{lemma:logreldeter} enables us to show the following
improved inversion lemma for function types whose statement is free of
the relation $[[F]]$, analogous to the admissible rule \rref{I-PiAlt}.
\begin{lemma}[Pi Inversion Alt\footnote{\dotv{semtyping.v}{InterpExt\_Fun\_inv\_nopf}}]
  \label{lemma:piinvalt}
  Suppose $[[Interp I i Pi x : A . B S]]$, then there exists some $[[S0]]$
  such that the following constraints hold:
  \begin{itemize}
  \item $[[Interp I i A S0]]$
  \item $[[forall a, (# a in S0 implies (# exists S1 , Interp I i B {a
    / x}
    S1 #) #)]]$
  \item $[[S = { b | forall a, (# a in S0 , forall
    S1, (# Interp I i B {a / x} S1,  b a in S1 #) #) }]]$
  \end{itemize}
\end{lemma}
\begin{proof}
  Immediate from Lemmas~\ref{lemma:interpinv} and \ref{lemma:logreldeter}.
\end{proof}

@@

% The inductive
% definition of the logical relation in Figure~\ref{fig:logrel} requires
% the impredicativity of Coq's \texttt{Prop} sort since\scw{or ``so that''?} in \rref{I-Pi},
% the function $[[F]]$ can be later instantiated into the logical
% relation itself (e.g. in the proof of Lemma~\ref{lemma:piintroalt}).
In Coq, we
